---
title: | 
  | Assignment 1: The Causes of Economic Growth
  | Introduction to Applied Data Science
  | 2022-2023
author: |
  | Bas Machielsen
  | a.h.machielsen@uu.nl
date: April 2023
urlcolor: purple
linkcolor: purple
output: 
  pdf_document:
    includes:
      in_header: "preamble.tex"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

## Assignment 1: The Causes of Economic Growth

In this assignment, you will gather data from the _World Bank_ website, and augment it with data from the _Clio Infra_ website. Then, you will visualize these data using several graphs and tables, and test several hypotheses about the causes of economic growth. You will use this document to complete the code chunks which I have left unfinished to produce your own data analysis & visualization. 

To start with, please replace my name and e-mail address with yours. Then, remove the lines:

```
output: 
  pdf_document:
    includes:
      in_header: "preamble.tex"
```
from the document and replace them by:

```
output: pdf_document
```
Now, we're ready to start. For all of the code-related questions, please answer with code, and do not type (or copy) the answer from the console. Rather, let R _generate_ your answer. 

### 1. World Bank Data

The _World Bank_ collects and processes large amounts of data and generates them on the basis of economic models. These data and models have gradually been made available to the public in a way that encourages reuse. In particular, the databases of the World Bank are available on [https://data.worldbank.org/](https://data.worldbank.org/). It pays the effort to browse through the website, see if you can navigate your way through the website, and use the interface the World Bank provides you. 

Normally, if you were looking for data from the World Bank, you would go to the website, find your dataset, download it to `.xlsx` or any other format, and then import it into an R `data.frame` using `read_xslx(.)`, or something else. But, this reliance on manual downloads of spreadsheets of the data they are interested in can quickly become overwhelming, as the work is manual, time consuming, and not easily reproducible. 

Fortunately, however, there also exist an R package which allows you to browse swiftly through World Bank data, and easily download it as an R `data.frame`. You can get this package by installing:

```{r}
library(pacman)
p_load("wbstats", "tidyverse")
```

You can navigate the database by searching for terms:

```{r}
wbstats::wb_search("gdp per capita")
```

Afterwards, you can proceed to download data by executing `wb_data("indicator_id")`. You can then write this to a data.frame, and merge this data with other indicators to create a dataset. There exist many of these packages, and we will also use another today.

Apart from being easy to use, these packages also have another advantage: reproducibility. Collecting data by means of code allows other users to unambiguously reproduce your data collection process. 

Firstly, we will look for GDP growth data. 

**Question x**: pass a search query to `wb_search` for GDP growth data, and download the indicator for which the description matches "_GDP (current US$)_". The full description should read:

> GDP at purchaser's prices is the sum of gross value added by all resident producers in the economy plus any product taxes and minus any subsidies not included in the value of the products. It is calculated without making deductions for depreciation of fabricated assets or for depletion and degradation of natural resources. Data are in current U.S. dollars. Dollar figures for GDP are converted from domestic currencies using single year official exchange rates. For a few countries where the official exchange rate does not reflect the rate effectively applied to actual foreign exchange transactions, an alternative conversion factor is used.

```{r, cache = TRUE}
wbstats::wb_search("GDP")
gdp <- wbstats::wb_data('NY.GDP.MKTP.CD')
```

**Question**: Rename the variable `NY.GDP.MKTP.CD` to `gdp`. Remove the `NA` observations from the dataset. How many observations are there in the dataset in total?

```{r}
gdp <- gdp %>% 
  rename(gdp = `NY.GDP.MKTP.CD`) %>%
  filter(!is.na(gdp))

nrow(gdp)
```

**Question x**: How many observations per country are there? Show the first ten observations. 

```{r}
gdp %>%
  group_by(country) %>%
  summarize(count = n()) %>%
  head(10)
```
**Question x**: How many different years are there in the dataset? Put them in increasing order. 

```{r}
gdp %>%
  select(date) %>% 
  pull() %>%
  unique() %>% 
  sort()
```

**Question x**: For each country, what is the first and last year? Again show the first ten observations. 

```{r}
gdp %>%
  group_by(country) %>%
  summarize(first_year = min(date), last_year = max(date))
```

**Question**: Make a summary of the data, with the mean, median, sd, min and max values for `gdp`.

```{r}
gdp %>%
  summarize(mean = mean(gdp),
            median = median(gdp), 
            sd = sd(gdp), 
            min = min(gdp), 
            max = max(gdp))

```


**Question**: What country, in which year, had the lowest GDP? And the highest? Hint: use a function similar to `slice` from the `dplyr` package. 

```{r}
slice_min(gdp, gdp, n = 1)

slice_max(gdp, gdp, n = 1)
```

Next, we'll have a look at population data, which we can also retrieve from the World Bank database. 

```{r}
population <- wb_data("SP.POP.TOTL")
```

**Question**: Rename the population variable to `population` and overwrite this to memory.

```{r}
population <- population %>%
  rename(population = `SP.POP.TOTL`)
```

Finally, we'll merge `population` with `gdp` on the basis of _country_ and _year_. 

**Question**: use `left_join` to merge `gdp` (left data.frame) with `population` (right data frame). Check whether everything has gone correctly. Save this dataframe to memory as `gdp_pop`. Select only `country, date, isco3c.x, gdp` and `population`. Then, use `mutate()` to create a new variable, `gdp_cap = gdp / population`. Then, again apply `na.omit()`. Write this data.frame to memory to `data`. 

```{r}
data <- left_join(gdp, population, 
          by = c("country" = "country", 
                 "date" = "date")) %>%
  select(country, date, iso3c.x, gdp, population) %>%
  na.omit() %>%
  mutate(gdp_cap = gdp / population)

```

Now, let's collect a pre-made version of GDP per capita from the World Bank website. 

```{r get gdp_cap, eval = FALSE}
wb_search("gdp per capita")

alt_gdp_pc <- wb_data('NY.GDP.PCAP.CD') %>%
  filter(!is.na(`NY.GDP.PCAP.CD`))
```


**Question**: What is the correlation between these two variables? What does that mean?

```{r eval = FALSE}
cor(data$gdp_cap, alt_gdp_pc$NY.GDP.PCAP.CD)
```

### 2. World Tables: Capital Stock

Next, we'll proceed to find some potential determinants of GDP growth. One of the classical determinants of GDP per capita growth is the level of physical capital. Many models in macroeconomics explain economic well-being on account of the amount of capital in an economy. In particular, we'll look for a few measures from the Penn World Tables. This data has to be downloaded manually from [this website](https://www.rug.nl/ggdc/productivity/pwt/). You can download an Excel file. Make sure to put it in the right directory when reading it:

```{r}
p_load("readxl")

pwt <- readxl::read_excel('pwt100.xlsx', sheet = 3)
```


We are looking for the `cn` variable, which indicates _Capital stock at current PPPs (in mil. 2017US$)_. 

**Question**: Select the variables `countrycode, year`, and `cn`, and rewrite the dataframe to memory.

```{r}
pwt <- pwt %>%
  select(countrycode, year, cn)
```

Now, we're looking to merge the two data.frames `data` and `pwt`, on the basis of common country names and years. 

**Question:** Have a look at the two datasets below. On the basis of which two matched variables in both datasets do you have to perform the merge? 

```{r}
pwt

data
```

We can do this in various ways: we can perform `left_join`, `right_join`, `inner_join`, or `outer_join`, but we can also use the `merge` function. Although their arguments differ somewhat, the results should absolutely be the same provided you specify the `by` arguments correctly. In that case, you match one particular country-year from the left data.frame to the identical particular country-year from the right data-frame, and put all variables together.

**Question**: Do this. Perform a merge and save the resulting data.frame to `merged_data`. 

```{r}
merged_data <- merge(data, pwt, by.x = c("iso3c.x", "date"), by.y=c("countrycode", "year"))
```

Next, we want to average GDP per capita and Capital stock for each country present in the dataset. This can be done easily using the `mutate` function from the `tidyverse` package. However, we want to investigate _current_ GDP per capita, so we do not want to take too long an average.

Say we want to take an average over the years 2010-2020. 

**Question**: Finish the following code to compute the average of GDP per capita and Capital stock. Make sure you deal with NA's explicitly. Save this again to `merged_data` 

```{r}
merged_data <- merged_data %>%
  group_by(country, iso3c.x) %>%
  filter(between(date, 2010, 2020)) %>%
  summarize(
    # fill in your answer here
    avg_gdpc = mean(gdp_cap, na.rm = TRUE),
    avg_cn = mean(cn, na.rm = TRUE)
  )

```

### 3. Historical Antecedents: Clio Infra Data

Another important element of capital might be not only physical capital, but human capital! Instead of using _contemporary_ human capital to explain economic development in 2010-2020, we will use _historical_ human capital. For this, we can again use a package, called `Clio`, which aggregates various historical datasets. You can install and load it by:

```{r}
devtools::install_github("basm92/Clio")
library(Clio)
```

You can see what variables are available in this dataset by running:

```{r}
Clio::clio_overview() %>% head(10)
```

As a proxy to measure historical human capital, we'll use _Average Years of Education_ in 1930. We can download this by running:

```{r}
educ <- Clio::clio_get("Average Years of Education")
```

**Question**: Filter this dataset such that only observations from 1930 remain. 

```{r}
educ <- educ %>%
  filter(year == 1930)
```

**Question**: Merge this dataset to the `merged_data` set, so that the human capital proxy is added to the dataset. 

When discussing economic growth, some people also talk about a _reversal of fortune_ tendency: the countries that were relatively the most wealthy in or before the Middle Ages are among the poorest now, and vice versa. We also want to investigate such an hypothesis. In order to do so, we use a proxy for wealth from 1500, the urbanization ratio. This can also be downloaded from the Clio Infra database. 

**Question**: Now find and download Urbanization Ratio, filter the dataset such that only observations from 1500 remain, and save it to `urb`. 

```{r}
urb <- clio_get("Urbanization Ratio") %>%
  filter(year == 1500) 
```

Now, let's merge `educ` and `urb` together, and then subsequently merge this to the `merged_data` data.frame. 

```{r}
educ_urb <- merge(educ, urb, by = "ccode")
```

**Question**: Also remove `year.x`, `year.y` and `country.name.y` from the dataset.

```{r}
educ_urb <- educ_urb %>%
  select(-c(year.x, year.y, country.name.y))
```


Finally, we need to merge `educ_urb` to the `merged_data` data.frame. 

**Question**: Merge these two dataframes using `left_join`, with `merged_data` being the left data.frame. Rename `Average Years of Education` and `Urbanization Ratio` to `av_educ` and `urban` respectively. 

```{r}
final <- merged_data %>%
  left_join(educ_urb, 
            by = c("country" = "country.name.x")) %>%
  rename(av_educ = `Average Years of Education`, urban = `Urbanization Ratio`)
```


### 4. Summarizing and Analyzing the Data

**Question**: Create a descriptive statistics table using the variables we have obtained. In it, we want to display the mean, median, sd, min, max and number of observations. Hint: use the `modelsummary` package, and use the following syntax for the formula: `x1 + x2 + x3 + x4 ~ (mean + median + min + max)*Arguments(na.rm = TRUE) + N`. 

```{r}
library(modelsummary)

modelsummary::datasummary(data = final, avg_gdpc + avg_cn + av_educ + urban ~ 
                            (mean + median + min + max)*Arguments(na.rm = TRUE) + N)
```

Next, we would like to make a histogram of the four aforementioned variables, to investigate their distribution. To do so, we need data in so-called long-form:

```{r}
final_long <- final %>%
  pivot_longer(c(avg_gdpc, avg_cn, av_educ, urban), names_to = "variable", values_to = "value")
```

With this dataset, we can create a histogram with four facets:

**Question x**: Add the right geometry element to complete the histograms. 

```{r}
final_long %>%
  ggplot(aes(x = value, group = variable, fill = variable)) + 
  facet_wrap(~variable, scales = "free") +
  # fill in the right answer
  geom_histogram()
```

Next, we would like to make a plot of some of the data. In particular, we can use the `ggplot` library to create a scatterplot of urbanization in 1500 on the $x$ axis and $\log$ (GDP per capita) on the $y$ axis. That would allow us to get insight into the _reversal of fortune_ theory.

**Question**: Use `ggplot` to create a plot as described above. Hint: use `geom_point()` as the geometric attribute. Also try to see if you can display the country name corresponding to each dot. 

```{r warning = FALSE}
final %>%
  ggplot(aes(x = urban, y = log(avg_gdpc))) +
  geom_point() + 
  geom_text(aes(label=country), nudge_x = 0.01, nudge_y = 0.2)
```

**Question**: Do you interpret this as evidence for the _reversal of fortune_ theory? Why (not)? 

Secondly, we can plot GDP per capita ($y$-axis) against the capital stock ($x$-axis). In this case, we want to log-transform both of the variables. 

**Question**: Construct this plot. Also make sure to add a nice title and to change the axis titles appropriately. 

```{r warning = FALSE}
final %>%
  ggplot(aes(x=log(avg_cn), y = log(avg_gdpc))) + geom_point() +
  xlab("Log (Capital Stock)") +
  ylab("Log (GDP Per Capita)") +
  ggtitle("Relationship between Capital Stock and Economic Well-Being")
```

**Question**: Do you interpret this as evidence for the logic of most macroeconomic models, that a higher capital stock causes a higher level of income? Why (not)? 

Finally, we would also like to make a map displaying economic growth rates. In order to do so, we need the `sf` package, short for _Spatial Features_. This is an efficient format in which data used to construct maps are stored. We also need a couple of auxiliary packages:

```{r}
library(tidyverse); library(sf)
```

Possibly, we also have to install a couple of auxiliary packages:

```{r}
pacman::p_load("rgdal", "rgeos", "lwgeom")
```

Let us first find a map of the world: 

```{r}
library(maps)
world <- st_as_sf(map("world", plot = FALSE, fill = TRUE))
```

We converted the world map from the `maps` package to an `sf` data.frame.

```{r}
world
```

`world` is now a data.frame, countaining the names of countries and associated polygons. We can use this object to create a simple map:

```{r}
world_map <- ggplot(world) + 
  geom_sf(fill = "grey80", 
          col = "grey40", 
          lwd = 0.3)

world_map
```

It is also possible to change projections. [Here](https://www.nceas.ucsb.edu/sites/default/files/2020-04/OverviewCoordinateReferenceSystems.pdf) is a short primer on different projections. For you, this is not particularly relevant, but it allows you to pick a projection which you like. Here's an example:

```{r}
world_map +
  coord_sf(crs = "+proj=laea +y_0=0 +lon_0=0 +lat_0=0") +
  labs(subtitle = "Lambert Azimuthal Equal Area projection")
```

Should we want to zoom in on a particular part of the world, that is also possible. To do that, we can filter the dataframe based on many features, for example, on coordinates:

```{r}
sf_use_s2(FALSE)

world %>%
  st_crop(xmin = -22, 
          ymin = 35, 
          xmax = 40, 
          ymax = 70) %>% 
  ggplot() + geom_sf()
```


**Question x**: Take your `data.frame` world, and merge it with the `final` data.frame containing the GDP growth rates. Save this to a data.frame called `world_data`. 

```{r}
world_data <- world %>% 
  left_join(final, by = c("ID" = "country"))
```


Now, plot a map displaying `avg_gdpc` in different countries. Hint: use the `geom_sf(fill = avg_gdpc)` as geometric attribute. 

```{r}
world_data %>%
  ggplot() + geom_sf(aes(fill = log(avg_gdpc))) + scale_fill_viridis_b()
```

## The End
